Higher-Order unification for free!

### 2024-04-30 paper/main.tex
minor
some typos in conclusion

\noindent
These terms are problematic for the very same reason terms in \maybebeta are,
but they cannot be handled directly by the unification of the meta language, which
is only required to handle terms in \llambda. Still, for a $s \in \notllambda$
there may exist a substitution $\rho$ such that $\rho s \in \llambda$.

@@

Another work that is, somewhat surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.\\
~\\
\emph{The approach presented in this paper}
%provides a third option that 
addresses all the concerns
mentioned earlier. It takes advantage of the
unification capabilities of the meta language
at the price of handling problematic sub terms on
the side.
As a result %of this choice 
our encoding takes
advantage of indexing data structures and
mode analysis for clause filtering.
It is worth mentioning that we only replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable. % by the meta language logic programming engine.
Some preliminary benchmarks on the Stdpp
library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018} show
encouraging results: our type-class solver has some
overhead on small goals but is consistently faster than
the Coq one starting from goals made of XXX nodes.\\
\todo{D: how many?}
Our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment.
Moreover, the unification process we obtain is ready to take
advantage of potential improvements to the programming engine,
such as tabled search. Finally, our shallow encoding can
enable to apply forms of static analysis for
the meta language, such as determinacy, also to the object language.

We considered mechanizing our results with Abella~\cite{gacek2008abella},
especially in the early phase of this work, but unfortunately we could not.
The main reason is that the logic of Abella does not let one quantify over
predicates nor user the cut operator. Eliminating cut or
combinators like \elpiIn{map} is surely possible, but 
has a high cost in verbosity given how pervasively they are used.
Moreover, and more importantly, it creates a distance
between the verified and the actual code, which we found to
hinder our exploratory phase.

>>>>

\noindent
These terms are problematic for the very same reason terms in \maybebeta are,
but they cannot be handled directly by the unification of the meta language, which
is only required to handle terms in \llambda. Still, given $s \in \notllambda$
there may exist a substitution $\rho$ such that $\rho s \in \llambda$.

@@

Another work that is, somewhat surprisingly, only superficially related to ours is the
type-class engine built in Isabelle's meta language. In~\cite{wenzel97}
classes identify (simple) types, they are not higher order predicates
as in Coq, hence the solver does not require a higher-order unification
procedure.\\
~\\
\emph{The approach presented in this paper}
%provides a third option that 
addresses all the concerns
mentioned earlier. It takes advantage of the
unification capabilities of the meta language
at the price of handling problematic sub terms on
the side.
As a result %of this choice 
our encoding takes
advantage of indexing data structures and
mode analysis for clause filtering.
It is worth mentioning that we replace terms with variables only
when it is strictly needed, leaving the rest of the term structure intact
and hence indexable. % by the meta language logic programming engine.
Some preliminary benchmarks on the Stdpp
library~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018} show
encouraging results: our type-class solver has some
overhead on small goals but is consistently faster than
the Coq one starting from goals made of XXX nodes.
\todo{D: how many?}
Moreover our approach is flexible enough to accommodate different strategies
and heuristics to handle terms outside the pattern fragment.
Finally, our encoding is sufficiently shallow to: 1)
allow to lift forms of static analysis for
the meta language, such as determinacy, also to the object
language; 2) take advantage of future improvements to the
logic-programming engine of Elpi, such as tabled search.

We considered mechanizing our results with Abella~\cite{gacek2008abella},
especially in the early phase of this work, but unfortunately we could not.
The main reason is that the logic of Abella does not let one quantify over
predicates nor use the cut operator. Eliminating cut or
higher order
combinators like \elpiIn{map} or \elpiIn{forall2} is surely
possible, but 
has a high cost in verbosity given how pervasively they are used.
Moreover, and more importantly, it creates a distance
between the verified and the actual code, which we found to
hinder our exploration. We did, however, establish a
test suite with about a hundred cases.
It is available at \githubUrl.
